---
title: "Completing and submitting practice exercises"
editor: visual
toc: true
number-sections: true
---

## Access your exercise repository

### Register for an account and join the student team

-   You must [sign up for a GitHub user account](https://github.com/signup) and share your GitHub username with me before I can create an exercise repository for you.
-   Note that you must have [installed R and RStudio](https://posit.co/download/rstudio-desktop/) to complete exercises and [installed GitHub Desktop](https://desktop.github.com/) to submit exercises.
-   Soon after I get your username, you should expect an email asking you to accept an invitation to the "2023_Students" Team in the class [GitHub organization](https://github.com/bldgspatialdata).
-   This is a private repository just for *your* exercise assignments so the repository will be named: **\<your first name\>\_\<your last name\>** and can be accessed at: **https://github.com/bldgspatialdata/\<your first name\>\_\<your last name\>**.

::: {.callout-tip collapse="false" appearance="default" icon="true"}
### What is a repository?

A repository is a collection of files where changes can be tracked and managed using a version control system. Git is a version control system. GitHub is a web service that makes it easier to manage version-controlled repositories and collaborate with others. For more background, read [Ch. 1 Why Git? Why GitHub?](https://happygitwithr.com/big-picture) from Happy Git and GitHub for the useR by Jenny Bryan.
:::

### Set up GitHub credentials in RStudio

If you are using RStudio instead of GitHub Desktop to update your repository, you also need to set up some local credentials so GitHub knows that RStudio is allowed to make changes to your repositories. You

The `{usethis}` documentation [includes detailed article on connecting GitHub and R](https://usethis.r-lib.org/articles/git-credentials.html) but it is more detail than you may need. First, install the `{usethis}` package (if you don't have it installed already):

```{r}
# pak::pkg_install("usethis")
```

Now, there are four steps (the first step is optional):

1. [Set up two-factor authentication](https://usethis.r-lib.org/articles/git-credentials.html#turn-on-two-factor-authentication). This is optional but recommended.

2. Run `usethis::create_github_token()` to [get a personal access token](https://usethis.r-lib.org/articles/git-credentials.html#get-a-personal-access-token-pat) and copy the token to your clipboard. Make sure you are signed in to GitHub in your browser first! I also suggest setting the expiration to 90 days so you won't need to refresh the token until after the end of this course.

3. Run `gitcreds::gitcreds_set()` to [put your PAT in your Git credential store](https://usethis.r-lib.org/articles/git-credentials.html#put-your-pat-into-the-local-git-credential-store)

4. **Restart your session** and run `usethis::gh_token_help()` to double-check that the token is saved in the right place.


If you run into any trouble, you can also run `usethis::git_sitrep()` for more feedback on what is set up correctly and what (if anything) might need to be fixed.

## Download or update your exercise repository

### Downloading (a.k.a "cloning") your exercise repository

-   Clone the repository to your computer using GitHub Desktop or using the GitHub website. Review the GitHub documentation on [cloning a repository with GitHub Desktop](https://docs.github.com/en/desktop/contributing-and-collaborating-using-github-desktop/adding-and-cloning-repositories/cloning-and-forking-repositories-from-github-desktop) or [the GitHub website](https://docs.github.com/en/desktop/contributing-and-collaborating-using-github-desktop/adding-and-cloning-repositories/cloning-a-repository-from-github-to-github-desktop) for more detailed instructions.
-   Find and open folder where you downloaded the local copy of the exercise repository. By default, this folder should have the same name as the repository itself.
-   As long as you keep this folder on your computer, you should do not need to clone the repository again. If you delete your local files or switch computers, you need to repeat this step.

### Updating (a.k.a "syncing") your exercise repository {#sec-update-repo}

-   I will be adding new exercises to your repository each week using GitHub. You should "sync" the changes to the GitHub repository with your local copy of the project each week. Read the GitHub documentation on [Syncing your branch in GitHub Desktop](https://docs.github.com/en/desktop/contributing-and-collaborating-using-github-desktop/keeping-your-local-repository-in-sync-with-github/syncing-your-branch-in-github-desktop) for more detailed instructions.

## Complete the week's exercise

The process for completing will be mostly the same for each week of the class.

### Locating the week's exercise file

-   Locate the **\<your first name\>\_\<your last name\>/exercises** folder on your computer.
-   Open the `exercises.RProj` file in RStudio. Review the documentation on [managing projects in RStudio](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) or the R for Data Science [chapter on working with projects](https://r4ds.hadley.nz/workflow-scripts#projects) for more background on why projects are useful.
-   Using the "Files" tab pane, locate the `exercise_01.qmd` file (or the number exercise file for this week) and open it. The file extension (`.qmd`) is short for Quarto Markdown document.
-   If you can't find this week's exercise file, you may have forgotten to update the repository. Go back and check @sec-update-repo

::: {.callout-tip collapse="false" appearance="default" icon="true"}
### What is a Quarto document?

A Quarto document is a format that lets you mix "chunks" of R code with sections of text formatted using the Markdown formatting conventions. It is similar to an older format known as RMarkdown (`.Rmd`). [Quarto](https://quarto.org/) comes installed with RStudio but it is a separate command line application that can turn a `qmd` document into a standalone [website](https://quarto.org/docs/websites/), a [formatted PDF](https://quarto.org/docs/output-formats/pdf-basics.html), a [presentation](https://quarto.org/docs/presentations/), and more. This course website, for example, is built with Quarto.

You can see that a [front matter](https://quarto.org/docs/authoring/front-matter.html) section at the top of each document. This front matter (also known as [YAML](https://en.wikipedia.org/wiki/YAML)) controls the format, appearance, and content of the rendered document.

The [Hello, Quarto tutorial](https://quarto.org/docs/get-started/hello/rstudio.html) is a great place to learn how to render Quarto documents, run code chunks interactively, and format text using the markdown syntax.
:::

-   Complete the project by filling in the blank spaces, fixing any "broken" code in the document, or adding new code in the "code chunks" that look like this:

```{r}
1 + 1 # example code
```

### Running code and writing responses to the exercise prompts

-   Try executing the code in each chunk as you work through the practice exercise. The results from the code appear right below the chunk. If you get an error or a warning, try to use that information as a clue to figure out what you need to change.

-   When you are done with the exercise, there should be no more blank spaces like this one: `____`. The bonus exercises are optional but otherwise any code chunk that started empty should now have code inside. Some questions require short written responses which should be included above or below the code chunks—not inside!

-   If all of your code chunks run smoothly, you should be able to "render" the `.qmd` document to create a new HTML document. When I review completed assignments, trying to render the document will always be my first step.

-   Remember, you don't need to get everything right and it is OK if your document won't render. Just try your best with each part of the exercise.

:::{.callout-tip collapse=false appearance='default' icon=true}
## Tips for completing exercises
- Complete this exercise from top to bottom. Some later chunks of code may depend on an earlier chunk so avoid skipping around.
- If you get an error message, read it. If it doesn't make sense, try looking it up. R for Data Science reminds readers that [Google is your friend](https://r4ds.hadley.nz/workflow-help.html#google-is-your-friend).
- Read the documentation!
:::

## Committing your completed exercise to the repository

You may have saved your updated document to your computer but the last step is to save or "commit" those changes to the exercise repository.

"Committing" your changes and syncing the changes with the remote version of your repository on GitHub allows me to see your completed exercise and share feedback on your work.

### Commit changes with GitHub Desktop

The first and easiest way to commit changes is using GitHub Desktop.

- Review the GitHub documentation on [Committing and reviewing changes to your project in GitHub Desktop](https://docs.github.com/en/desktop/contributing-and-collaborating-using-github-desktop/making-changes-in-a-branch/committing-and-reviewing-changes-to-your-project-in-github-desktop) for more detailed instructions.

### Commit changes with RStudio

The second way to commit changes is using the Git tab within RStudio. The interface for Git within RStudio is a little less friendly but is a convenient option since you don't need to leave RStudio to commit your changes.

Here is the step-by-step process (adapted from [Connect RStudio to Git and GitHub](https://happygitwithr.com/rstudio-git-github) in Happy Git and GitHub for the useR):

-   Click the "Git" tab in upper right pane.
-   Check "Staged" box for `exercise_01.qmd`.
-   If you're not already in the Git pop-up, click "Commit".
-   Type a message in "Commit message", such as "Commit from RStudio".
-   Click "Commit".
-   Finally, click the green "Push" button to send your local changes to GitHub.

::: {.callout-note collapse="false" appearance="default" icon="true"}
### When should you commit your changes?

You don't need to wait until you are 100% finished with the exercise to commit your changes. You may want to commit the changes every time you work on the exercise. Please make sure to change the "status" for the exercise from "Available" to "Complete" when your exercise is complete and ready for evaluation. I will typically evaluate the last week's exercises on the Tuesday before each class session.
:::
