---
title: "Coding basics"
execute: 
  error: true
---

## Week 1

### Packages in R

A container for:

-   functions
-   data
-   documentation

`{tidyverse}` is a *metapackage*—or a package that loads a set of other packages.

#### Installing packages

Use `install.packages()` to install a package from [CRAN](https://cran.r-project.org/) (Comprehensive R Archive Network):

```{r}
install.packages("sf")
```

But, remember, this does not work:

```{r}
install.packages(sf)
```

#### Loading packages

Use `library()` to load a package into your environment:

```{r}
library(sf)
```

::: {.callout-note collapse="true"}
#### Why can we use both `library(sf)` and `library("sf")`?

The `package` argument can be the name of a package: as a **name**, **literal character string**, or **a character string**.
:::

#### base functions and packages are always available

The `{base}` package for R is one of fifteen "base packages" that are always available. Functions like `sum()`, `max()`, and `paste0()` are all examples of "base functions".

### Functions in R

Functions take inputs (known as arguments or parameters) and return outputs.

A container for:

-   Logic (a.k.a. an algorithm)
-   Math
-   More functions
-   Data *or* connections to data sources

------------------------------------------------------------------------

### Take a look at `st_read()`

```{r}

```


------------------------------------------------------------------------

Some functions require a specific type of input:

```{r}
sum(1, 2, 3)

sum("A", "B", "C")
```

------------------------------------------------------------------------

But some functions are more flexible:

```{r}
paste0(1, 2, 3)

paste0("A", "B", "C")
```

------------------------------------------------------------------------

Order of execution matters!

```{r}
var + 2
var <- 1
```
