---
title: "Weekly Updates"
editor: visual
date: 2024-09-04
number-sections: false
---

## Week 9

- Exercise 5 due next Friday, Nov. 1
- Office hours on Tuesday, Oct. 29
- Final project proposal due Wednesday, Nov. 13
- Lessons from McConchie and Boeing
- Open Street Map Lecture & Practice

### Lessons McConchie (2016)

McConchie introduces the concept of **map-gardening:** "the editing tasks that keep OSM going... the things that happen after that fun trailblazing phase of mapping all the streets in your neighborhood."

McConchie asks: "How do we make sure that OSM is a healthy community that has gardening, has people who enjoy maintenance?

---

{{< video https://www.youtube.com/watch?v=7sC83j6vzjo >}}

### Lessons from Boeing (2020)

Why did Boeing make OSMnx open source? Making the tool open source:

-   "makes empirical work easier to review and reproduce."
-   "allows anyone else to contribute to the tool’s ongoing development."
-   "empowers others working in urban science and planning to advance their empirical research on real-world spatial networks with a reusable, accessible, theoretically-sound tool."

---

Why did Boeing use OpenStreetMap data?

-   Google Maps data is unavailable (or un-affordable)
-   TIGER/Line roads shapefiles don't include topological details
-   Open Street Map is free and has global coverage

## Week 4

-   Exercise 2 solutions
-   Data transformation with dplyr
-   Parsons Problems

### Parsons Problems with {dplyr}

------------------------------------------------------------------------

```{r}
#| eval: false
#| code-line-numbers: true
  wind > 130
storms |>
filter(
library(dplyr)
  year == 2010,
)
```

```{r}
#| echo: false
#| output: false
library(dplyr)

storms |>
  filter(
    year == 2010,
    wind > 130
  )
```

------------------------------------------------------------------------

```{r}
#| eval: false
#| code-line-numbers: true
slice_head(n = 10)
library(dplyr)
arrange(desc(wind_load)) |>
distinct(year, name, .keep_all = TRUE) |>
storms |>
mutate(wind_load = pressure * wind^2) |>
```

```{r}
#| echo: false
#| output: false
library(dplyr)
storms |>
  mutate(wind_load = pressure * wind^2) |> 
  arrange(desc(wind_load)) |> 
  distinct(year, name, .keep_all = TRUE) |> 
  slice_head(n = 10)
```

------------------------------------------------------------------------

```{r}
#| eval: false
#| code-line-numbers: true
ggplot(aes(x = category, y = avg_wind)) +
storms |>
summarise(avg_wind = mean(wind)) |>
library(tidyverse)
filter(!is.na(category)) |>
geom_point()
group_by(category) |>
```

```{r}
#| echo: false
#| output: false
library(tidyverse)

storms |>
  filter(!is.na(category)) |>
  group_by(category) |>
  summarise(avg_wind = mean(wind)) |>
  ggplot(aes(x = category, y = avg_wind)) +
  geom_point()
```

------------------------------------------------------------------------

## Week 3

-   Exercise submission process
-   Exercise 1 solutions
-   Mapping with ggplot2
-   Data transformation with dplyr
-   Questions

## Week 2

-   Syllabus updates
-   Cheat sheets
-   Check-in on exercise 1 and GitHub Classroom
-   Week 2 in-class quiz
-   Visualizing spatial data with ggplot2

### Syllabus updates

-   In-class exercises
-   No class on November 27

## Cheat sheets

-   RStudio
-   sf
-   ggplot2
