---
title: "Creating and manipulating attributes for spatial data"
order: 4
editor: visual
exec:
  eval: false
---

## Setup

Today, we are going to use the `{tidyverse}` along with `{sf}` and two related packages: `{lwgeom}` and `{units}`.

```{r}
library(tidyverse)
library(sf)
library(lwgeom)
library(units)
```

------------------------------------------------------------------------

## Getting started

For this session, we also need some data to look at. We are going to load data from the U.S. Census Bureau using the `{tigris}` package:

```{r}
library(tigris)
options(tigris_use_cache = TRUE)
```

------------------------------------------------------------------------

Use `states()` to load data on U.S. states and `primary_roads()` to load data on U.S. highways:

```{r}
us_states <- states()

us_highways <- primary_roads()
```

```{r}
#| echo: false
compare_sf <- function(x, y, colors = c("orange", "purple")) {
ggplot() +
  geom_sf(data = x, color = colors[[1]]) +
  geom_sf(data = y, color = colors[[2]]) +
  theme_minimal()
}
```

## Things to remember about spatial data

-   Fields
-   Objects

Check out the [Wikipedia article on data models in GIS](https://en.wikipedia.org/wiki/Data_model_(GIS)) for more background on this topic.

## Things to remember about `sf` and `sfc` objects

A `sf` object is a data frame with a `sfc` list-column.

```{r}
us_states$geometry
```

## Things to remember about `sf` and `sfc` objects

-   a `sf` object is a data frame with a `sfc` list-column
-   a `sf` object has a `sf_column` attribute (it isn't always named geometry---use `attributes()` to take a look)

------------------------------------------------------------------------

```{r}
attributes(us_states)
```

## Things to remember about `sf` and `sfc` objects

-   a `sf` object is a data frame with a `sfc` list-column
-   a `sf` object has a `sf_column` attribute (it isn't always named geometry---use `attributes()` to take a look)
-   `sf` and `sfc` objects use a coordinate reference system

------------------------------------------------------------------------

```{r}
st_crs(us_states)

st_crs(us_states$geometry)
```

## Things to remember about `sf` and `sfc` objects

-   a `sf` object is a data frame with a `sfc` list-column
-   a `sf` object has a `sf_column` attribute (it isn't always named geometry---use `attributes()` to take a look)
-   `sf` and `sfc` objects use a coordinate reference system
-   working with `sf` objects is *slower* than working with data frames---so drop the geometry if you don't need it

------------------------------------------------------------------------

```{r}
st_drop_geometry(us_states)
```

## Things to remember about coordinate reference systems

-   objects *must* share the same coordinate reference system if you are using them together
-   coordinate reference systems are stored as attributes for `sfc` and `sf` objects (`sfg` objects don't have a CRS)
-   coordinate reference systems have *units*
-   geographic and projected coordinate reference systems are *not* the same
-   coordinate reference systems can be missing and they can be wrong

------------------------------------------------------------------------

```{r}
#| eval: false
# bad: assigning a CRS that doesn't match the geometry
st_crs(us_states) <- 3857

# good: using st_transform to convert the geometry to a new CRS
st_transform(us_states, crs = 3857)
```

------------------------------------------------------------------------

Get the objects into the same CRS before we continue:

```{r}
us_states <- st_transform(us_states, crs = 3857)

us_highways <- st_transform(us_highways, crs = 3857)
```

## Things to think about

-   `sf` objects are *not* the only way to represent spatial data in R

-   If you are working with more than one `sf` or `sfc`, the objects must use the *same* coordinate reference system to use them together.

```{r}
# state
```

## What is an "attribute" in GIS?

Within a GIS desktop application, an **attribute** may be known as a **field**.

When we talk about tidy data frames, an attribute is equivalent to a **variable** which is represented as a **column** in a data frame.

::: notes
https://support.esri.com/en-us/gis-dictionary/attribute

-   \[data models\] Nonspatial information about a geographic feature in a GIS, usually stored in a table and linked to the feature by a unique identifier. For example, attributes of a river might include its name, length, and sediment load at a gauging station.
-   \[data models\] In raster datasets, information associated with each unique value of a raster cell.
-   \[graphics (map display)\] Information that specifies how features are displayed and labeled on a map; for example, the graphic attributes of a river might include line thickness, line length, color, and font for labeling.
:::

------------------------------------------------------------------------

**But!** Feature geometry is also stored as a column.

A single `sfc` object (or list-column) can contain more than one feature (shapefiles can't do this BTW!) using MULTIPOINT, MULTILINESTRING, MULTIPOLYGON, or GEOMETRYCOLLECTION geometry types.

------------------------------------------------------------------------

**And!** Objects in R can also have attributes and these are *not* the same thing as attributes for spatial data. Try not to get them mixed up.

## What is an attribute?

So (to recap) an *attribute* in GIS can also be called a...

-   ...**field** in a desktop GIS application
-   ...**variable** in tidy data
-   ...**column** in a data frame

## What types of attributes exist?

Attributes are "properties of features ('things') that do not describe the feature's geometry".

Here are the attributes for `us_states`:

```{r}
glimpse(st_drop_geometry(us_states))
```

------------------------------------------------------------------------

Attributes can have:

-   **point support:** the value applies to every point *individually*, or
-   **block support:** the value is a summary for *all* points in the geometry

------------------------------------------------------------------------

The relationship between attributes and geometry can be described as:

-   **constant:** the value is valid everywhere in or over the geometry
-   **aggregate:** the value is associated with the entire geometry

------------------------------------------------------------------------

What is an example of an attribute with a constant relationship?

```{r}
glimpse(us_states)
```

::: notes
-   NAME, REGION, or DIVISION: Yes - every part of a state can still use the same state name
-   ALAND: No - this is an aggregate value
-   INTPTLAT: No - this is a summary value
:::

------------------------------------------------------------------------

What is an example with an aggregate relationship?

```{r}
glimpse(us_states)
```

::: notes
-   NAME, REGION, or DIVISION: Yes - every part of a state can still use the same state name
-   ALAND or AWATER: Yes - this is an aggregate value - if the shape of the state changed, the land area would change
-   INTPTLAT or INTPTLON: Yes - this is a summary value - if the shape of the state changed, the interior latitude/longitude point would change

Illustrates the point: "A typical property of such variables is that associated geometries come for instance from legislation, observation devices or analysis choices, but not intrinsically from the observed variable."
:::

------------------------------------------------------------------------

Attributes can be:

-   extensive: corresponding to amounts, associated with a physical size (length, area, volume, counts of items)
-   intensive:

## What is an "attribute domain"?

::: notes
-   \[data structures\] In a geodatabase, a mechanism for enforcing data integrity. Attribute domains define what values are allowed in a field in a feature class or nonspatial attribute table. If the features or nonspatial objects have been grouped into subtypes, different attribute domains can be assigned to each of the subtypes.
:::

## What are we working with?

When we do data analysis using `{dplyr}`, there are three types of functions we use most often:

-   Boolean operators or predicates
-   Window or vector functions
-   Summary or analysis functions

There are similar

## Predicate functions for geometries with `{sf}`

`{sf}` includes "vectorized" logical operators or tests that work with geometry including:

-   `st_is()`
-   `st_is_valid()`
-   `st_is_empty()`

------------------------------------------------------------------------

```{r}
st_is(us_states, "POLYGON")

st_is(us_states, "MULTIPOLYGON")
```

## Predicate functions for geometries with `{sf}`

`{sf}` also includes more than a dozen predicate functions for working with *pairs* of simple geometries including:

-   `st_intersects`
-   `st_disjoint`
-   `st_contains`
-   `st_covers`
-   `st_is_within_distance`

## Using predicate functions for spatial joins and filters

`st_filter()` and `st_join()` are two functions that rely on these predicate functions to work.

------------------------------------------------------------------------

```{r}
maryland <- filter(us_states, NAME == "Maryland")

st_filter(us_highways, maryland)
```

------------------------------------------------------------------------

```{r}
st_filter(us_states, maryland, .predicate = st_disjoint)
```

## Creating new variables with geometry

-   Measuring feature geometries
-   Comparing feature geometries
-   Joining data based on feature geometries

```{r}
maryland <- filter(us_states, NAME == "Maryland")
```

## Measuring geometries with `{sf}`

## Measuring geometries with `{sf}`

`{sf}` includes a few different functions for measuring geometries:

-   `st_area()` (only works with POLYGON and MULTIPOLYGON geometries)
-   `st_length()` (only wrks with LINESTRING and MULTILINSTRING geometries)
-   `st_distance()` (requires a pair of objects)

## Measuring geometries with `{sf}`

All of these functions are *vectorized* meaning that they can operate independently on each feature in a `sf` or `sfc` object.

They support both `sf` inputs (data frames) *and* `sfc` inputs (lists)---but they always return a vector:

```{r}
st_area(maryland)

st_area(maryland$geometry)
```

## Measuring geometries with `{sf}`

But, remember, `dplyr::mutate()` is designed to work with vectorized functions so you can do us a measurement function inside `mutate()`:

```{r}
maryland |> 
  mutate(
    area = st_area(geometry)
  )
```

## Measuring geometries with `{sf}`

You can even work with multiple geometries using this same approach:

```{r}
us_states |> 
  mutate(
    area = st_area(geometry),
    distance_to_maryland = st_distance(geometry, maryland)
  )
```

This works to aggregate features by division:

```{r}
us_states |> 
  group_by(DIVISION) |> 
  summarise(
    n_states = n_distinct(NAME)
  ) 
```

------------------------------------------------------------------------

```{r}
us_states |> 
  group_by(DIVISION) |> 
  summarise(
    n_states = n_distinct(NAME)
  ) |> 
  ggplot() +
  geom_sf(aes(fill = DIVISION)) +
  theme_minimal()
```

------------------------------------------------------------------------

This doesn't entirely work (yet):

```{r}
us_states |> 
  group_by(DIVISION) |> 
  summarise(
    n_states = n_distinct(NAME)#,
    # .by = DIVISION
  )
```

------------------------------------------------------------------------

But *this* (explicitly unioning the geometry) does work:

```{r}
us_states |> 
  summarise(
    n_states = n_distinct(NAME),
    geometry = st_union(geometry),
    .by = DIVISION
  )
```

------------------------------------------------------------------------

```{r}
us_states |> 
  group_by(DIVISION) |> 
  summarise(
    n_states = n_distinct(NAME),
    geometry = st_union(geometry)
  ) |> 
  st_as_sf() |> 
  ggplot() +
  geom_sf(aes(fill = DIVISION)) +
  theme_minimal()
```

## Measuring geometries with `{sf}`

Not all functions work with all geometry types!

-   `st_area()` only works with POLYGON and MULTIPOLYGON geometries
-   `st_length()` only works with LINESTRING and MULTILINSTRING geometries

## Measuring geometries with `{sf}`

-   `st_distance()` requires at least two objects

## Measuring geometries with `{lwgeom}` and `{geosphere}`

-   `lwgeom::st_perimeter()`
-   `geosphere::bearing()`

------------------------------------------------------------------------

```{r}
us_states |> 
  group_by(DIVISION) |> 
  summarise(
    n_states = n_distinct(NAME)
  )
```
