---
title: "Spring 2026 Weekly Updates"
editor: visual
number-sections: false
scrollable: true
---

```{r}
#| echo: false
library(sf, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(spData)
```

## Week 3

- Week 3 Questions
- [Session 2 Quiz](https://docs.google.com/forms/d/e/1FAIpQLScFfOG6aFaMv-pogR2wZHoSYnfhmaemRwHyytWCZPpSJV6dog/viewform?usp=header)
- [Visualizing spatial data with {ggplot2}](/slides/data-visualization.qmd)

### Questions

What are some effective ways to familiarize yourself with the language of different packages without rote memorization of their functions? —Brian

- [Cheatsheets](https://posit.co/resources/cheatsheets/), e.g. [RStudio](https://rstudio.github.io/cheatsheets/html/rstudio-ide.html), [ggplot2](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
- Community resources, e.g. [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
- Naming conventions, e.g. `geom_<type of geometry to show on plot>`

----

Is there an easy way to plot summary statistics (e.g. mean, min, max)? —Lauren

```{r}
summary(spData::us_states_df)
```

----

Is there an easy way to plot summary statistics (e.g. mean, min, max)? —Lauren

```{r}
skimr::skim(spData::us_states_df)
```

----

Why are some CRS values `NA` while some are specific datum (WGS84, etc)? —Nhi

```{r}
x <- st_point(c(1, 0))

st_crs(x)

st_crs(spData::us_states)
```

----

Why do some of the exercises start with the library function and not have you initially install a library? —Kyle

```{r}
# install.packages("tidyverse")
```

----

What actually is an observation and how is it different than a variable? —Dillon

A variable is something you can *measure*. An observation is a set of *measurements*.

----
