---
title: "Spring 2026 Weekly Updates"
editor: visual
number-sections: false
scrollable: true
---

```{r}
#| echo: false
library(sf, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(spData)
```

## Week 6

-   Schedule adjustment for 4/1
-   Reminders
    -   Exercise 3 due today
    -   Exercise 4 due Wednesday, 3/11
    -   Exercise 5 coming soon
    -   Exercise 1, 2, or 3 questions?
-   Final Project Timeline
    -   Project proposal - due Wednesday, 3/25
    -   Final project presentations - due Monday, 5/4
    -   Final project repository - due Friday, 5/15
-   Questions
-   Recap: [Creating and manipulating attributes for spatial data](/slides/spatial-data-attributes.qmd)
-   Lecture: [Working with spatial and geometric operations in {sf}](/slides/spatial-data-attributes.qmd)
-   Review [two-table verbs](https://dplyr.tidyverse.org/articles/two-table.html) with dplyr

### Questions

How do you decide the “right” structure for a dataset when it can be organized in more than one tidy way? - Vrinda

Typically, whatever format allows you to efficiently complete the necessary analysis and produce the expected outputs is the "right" structure.

----

How much does it matter when making the decision on what function to use when smoothing? - Dillon

Check out the [smoothr](https://strimas.com/smoothr/) documentation for more details.

You can also use `sf::st_simplify()` or `rmapshaper::ms_simplify()` for make less "smooth" lines and polygons.

----

Which carrier has the worst average delays? Check [R for Data Science (2e) - Solutions to Exercises](https://mine-cetinkaya-rundel.github.io/r4ds-solutions/) for tips.

```{r}
nycflights13::flights |>
  group_by(carrier) |>
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) |>
  slice_max(order_by = avg_delay, n = 1)
```

::: footer
There are also solutions available for [Geocomputation with R](https://r.geocompx.org/solutions/).
:::

----

Can you disentangle the effects of bad airports vs. bad carriers? (via [Solutions Manual: R for Data Science (2e)](https://aditya-dahiya.github.io/RfDS2solutions/))

```{r}
nycflights13::flights |>
  group_by(dest, carrier) |>
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) |>
  # taking the highest average delay flight at each airport
  slice_max(order_by = avg_delay, n = 1) |>
  ungroup() |>
  # for each airline, summarize the number of airports where it is
  # the most delayed airline
  summarise(n = n(), .by = carrier) |>
  slice_head(n = 5) |>
  arrange(desc(n)) |>
  rename(Carrier = carrier, `Number of Airports` = n) |>
  gt::gt()
```

----

How do you choose when to use `st_intersection()` vs. `st_join()` when looking at relationships between layers? - Liam

```{r}
nc <- st_read(system.file("shape/nc.shp", package = "sf"))

nc_points <- st_as_sf(st_sample(nc, 250))

st_join(nc_points, nc[1, ])

st_intersection(nc_points, nc[1, ])
```

## Week 5

-   Reminder
    -   Exercises 2 due Friday, 2/27
    -   Exercise 3 due Wednesday, 3/4
-   Interesting and Difficult Things!
-   Week 5 Questions
-   Finish review of [Transforming data with `{dplyr}`](/slides/data-transformation.qmd)
-   [Creating and manipulating attributes for spatial data](/slides/spatial-data-attributes.qmd)

### Interesting Things

-   "honestly was just surprised how easy it is to perform geometry operations"
-   "we can use distance based joins on two datasets that are meaningfully related even when they don’t intersect"

### Difficult Things

-   "learning how to do the same thing in multiple different ways"
-   "understanding the different binary operators... st_intersects and st_disjoint were intuitive but others like st_covered_by were less clear"
-   "finding the time to do the readings"

### Questions

"Have you ever had to use DE-9IM strings, will I ever have to use them, can they be practically effectively used by people who are not deep into the lore??" - Lauren

------------------------------------------------------------------------

"Do unary and binary geometry operations change the input file or allow you to create a unique output file" - Connor

## Week 4

-   Updates and reminders
    -   Common issues with syntax
    -   Exercise how-to with Quarto
    -   Exercises 2 and 3 due Friday, 2/27 and Wednesday 3/4 (links coming soon)
    -   Office Hours on Friday, 2/20 at 12:00 pm
-   Week 4 Questions
-   Finish review of data visualization example with ggplot2
-   [Transforming data with `{dplyr}`](/slides/data-transformation.qmd)

### Questions

When subseting with the `[` operator, why do you need a comma and space at the end? e.g. `world[world$area_km2 < 10000, ]` —Chase

Use `?[` to take a look at the documentation. When using the `[` operator to subset a data frame (or an `sf` object), the first value is the row index and the second is the column index.

------------------------------------------------------------------------

```{r}
# Select second row
us_states[2, ]

# Select second column
us_states[, 2]
```

------------------------------------------------------------------------

Do you always use `sf` objects when working with spatial data? Or do you switch between spatial and non-spatial formats? —Liam

Yes. Always drop the geometry using `sf::st_drop_geometry()` if you don't need it in your output!

```{r}
#| code-fold: true
#| code-summary: "Summarising a data frame vs. sf object with `bench::mark`"

storms_sf <- sf::st_as_sf(storms, coords = c("long", "lat"), crs = 4326)

comparison <- bench::mark(
  df = storms |>
    dplyr::summarise(
      mean_wind = mean(wind),
      .by = year
    ),
  sf = storms_sf |>
    dplyr::group_by(year) |>
    dplyr::summarise(
      mean_wind = mean(wind)
    ),
  check = FALSE
)
```

------------------------------------------------------------------------

```{r}
#| echo: false
plot(comparison)
```

------------------------------------------------------------------------

Does it matter the order that you specify parameters for a `ggplot`? —Lauren

Consistent [code style](https://r4ds.hadley.nz/workflow-style.html) improves the readability of your code and reduces risk of errors *but* `{ggplot2}` supports a flexible approach.

------------------------------------------------------------------------

This works...

```{r}
ggplot(data = us_states) +
  geom_sf(aes(fill = total_pop_15))
```

------------------------------------------------------------------------

...and this works...

```{r}
ggplot(data = us_states) +
  aes(fill = total_pop_15) +
  geom_sf()
```

------------------------------------------------------------------------

...and this works!

```{r}
ggplot() +
  geom_sf(data = us_states, aes(fill = total_pop_15))
```

------------------------------------------------------------------------

But... this does not work! Do you know why?

```{r}
ggplot() +
  geom_sf(us_states, aes(fill = total_pop_15))
```

### Finish review

Download the example script to review:

```{r}
#| eval: false
usethis::use_github_file(
  repo = "bldgspatialdata/website",
  path = "examples/data-visualization.R"
)
```

## Week 3

-   Week 3 Questions
-   [Session 2 Quiz](https://docs.google.com/forms/d/e/1FAIpQLScFfOG6aFaMv-pogR2wZHoSYnfhmaemRwHyytWCZPpSJV6dog/viewform?usp=header)
-   [Visualizing spatial data with {ggplot2}](/slides/data-visualization.qmd)

### Questions

What are some effective ways to familiarize yourself with the language of different packages without rote memorization of their functions? —Brian

-   [Cheatsheets](https://posit.co/resources/cheatsheets/), e.g. [RStudio](https://rstudio.github.io/cheatsheets/html/rstudio-ide.html), [ggplot2](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
-   Community resources, e.g. [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
-   Naming conventions, e.g. `geom_<type of geometry to show on plot>`

------------------------------------------------------------------------

Is there an easy way to plot summary statistics (e.g. mean, min, max)? —Lauren

```{r}
summary(spData::us_states_df)
```

------------------------------------------------------------------------

Is there an easy way to plot summary statistics (e.g. mean, min, max)? —Lauren

```{r}
skimr::skim(spData::us_states_df)
```

------------------------------------------------------------------------

Why are some CRS values `NA` while some are specific datum (WGS84, etc)? —Nhi

```{r}
x <- st_point(c(1, 0))

st_crs(x)

st_crs(spData::us_states)
```

------------------------------------------------------------------------

Why do some of the exercises start with the library function and not have you initially install a library? —Kyle

```{r}
# install.packages("tidyverse")
```

------------------------------------------------------------------------

What actually is an observation and how is it different than a variable? —Dillon

A variable is something you can *measure*. An observation is a set of *measurements*.

------------------------------------------------------------------------
