---
title: "Spring 2026 Weekly Updates"
editor: visual
number-sections: false
scrollable: true
---

```{r}
#| echo: false
library(sf, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(spData)
```

## Week 12

- **Discuss**: What to do with week 14?

- Final project

  - **Proposal:** due 4/11

  - **Presentation**: due 5/4

  - **Final repository**: due 5/15

- Questions

- [Reading and writing spatial data](/slides/spatial-data-io.qmd) (and using web services)

### Questions

**Are the cartographic boundary version of tiger data less accurate? Why would someone not use them?**

[Cartographic boundary files](https://www.census.gov/geographies/mapping-files/time-series/geo/cartographic-boundary.html) are designed for *display* purposes. More detail may not be legible for a larger scale map—so cartographic files are scale-specific ("500k", "5m", or "20m"). [Natural Earth](https://www.naturalearthdata.com/) data is also provided at multiple scales for cartographic purposes.

------------------------------------------------------------------------

**Is it better to clean data in Excel/Sheets first or bring messy data into R and clean it there?**

It is OK to use a spreadsheet when you are getting started—but switching to a R *only* workflow has advantages. "Cleaning" the data is a *transformative* process so always make sure to make a copy before modifying a source file.

------------------------------------------------------------------------

**Since using Excel or CSV for spatial data is risky in professional workflows. Since these formats are so commonplace in the work place what can we to help prevent issues like loss of CRS, missing metadata, and data errors?**

Using `{openxlsx2}` you can set up Excel workbooks with protected sheets or ranges (see how-to in [Excel](https://support.microsoft.com/en-au/office/protect-a-worksheet-3179efdb-1285-4d49-a9c3-f4ca36276de6) or [Sheets](https://support.google.com/docs/answer/1218656?hl=en&co=GENIE.Platform%3DDesktop)) or validation. You can use `{labelled}` to create a data dictionary and include it with your workbook or sheet.

Read more on [Data Organization in Spreadsheets](https://www.tandfonline.com/doi/full/10.1080/00031305.2017.1375989).

------------------------------------------------------------------------

**What is stopping ERSI Shapefile from being improved so that the file size can be larger and column names can be less restricted?**

Systems and software are designed to support specific file format standards. Standards can change—adding or removing options for people building tools around the standard—but then software must also change to support it.

------------------------------------------------------------------------

What does "driver" mean in the context of `sf::st_drivers()`?

GDAL [defines a driver](https://gdal.org/en/stable/glossary.html#term-Driver) as: "A software component that enables reading, writing, and processing of specific raster or vector data formats."

## Week 11

### Questions

**When using OSM how can we be sure that we are using data that is accurate and up to date?**

Comparing to a reference data or using metadata from OSM itself are two common approaches.

A [2021 study](https://www.sciencedirect.com/science/article/pii/S1195103624005573) found significant associations between road network completeness, positional accuracy, and attribute accuracy and a set of five indicators: "population, average income, density of OSM roads, density of OSM buildings, and number of points of interest (POI)."

------------------------------------------------------------------------

**When writing functions that take a dataset as input, should we assume it will be a tibble or a data frame? Does it affect how we write the function? When should we use them (tibble vs df)?**

Check out the [tibble vignette](https://tibble.tidyverse.org/articles/tibble.html) for more details. Key differences for tibbles include not allowing partial matching of column names, using stricter rules for column names, and not allowing row names.

------------------------------------------------------------------------

**Is it possible to prepare a complete thesis using Quarto and generate the final document in PDF or Word format with all required sections and formatting?**

Yes! Quarto extensions include [specific journal formats](https://quarto.org/docs/journals/formats.html) and, with version 1.9, Quarto added support for [Typst format book projects](https://quarto.org/docs/blog/posts/2026-03-31-typst-books-and-more/#typst-books) including options to have chapters and appendices. For one example see [How I’m Writing My Dissertation in Quarto](https://carlcolglazier.com/notes/disseration-quarto/) by Carl Colglazier.

------------------------------------------------------------------------

**tigris and tidycensus questions**

- Is there any reason for the specific years available in tigris? A lot of functions such as divisions(), places(), or pumas() range from 2013-, 2012-, or 2011-. Why do so many go back to specifically the early 2010s?

- How do you know which geographic level (tract vs. block group vs. county) is the best to use for an analysis? Does it really change your results that much?

- How does areal interpolation work to account for disparate zonal configurations across years?

- In my job I often produce maps using data from the ACS. Given that these maps often have large margins of error how should I go about creating choropleth maps that will be used by other professionals with little to no understanding of field GIS?

## Week 9

- Questions
- Geocoding with [tidygeocoder](https://jessecambon.github.io/tidygeocoder/) or [arcgisgeocode](https://r.esri.com/arcgisgeocode/)
- Writing functions in R
- Creating projects with Quarto
- Exploratory data analysis

### Questions

- What does the "loess" method mean in the [`geom_smooth`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) function? "loess" is for LOESS (locally estimated scatterplot smoothing) - a [local regression](https://en.wikipedia.org/wiki/Local_regression) method.

------------------------------------------------------------------------

- Is there a rule of thumb of when to create a function or is it mostly just "you know you need one when you think you do"?
- How do I decide when it’s worth turning code into a function instead of just keeping it in-line?
- In data science workflows, how do we best balance the Don't Repeat Yourself principle and the use of existing packages with the educational value of writing algorithms from first principles?

## Week 6

- Schedule adjustment for 4/1
- Reminders
  - Exercise 3 due today
  - Exercise 4 due Wednesday, 3/11
  - Exercise 5 coming soon
  - Exercise 1, 2, or 3 questions?
- Final Project Timeline
  - Project proposal - due Wednesday, 3/25
  - Final project presentations - due Monday, 5/4
  - Final project repository - due Friday, 5/15
- Questions
- Recap: [Creating and manipulating attributes for spatial data](/slides/spatial-data-attributes.qmd)
- Lecture: [Working with spatial and geometric operations in {sf}](/slides/spatial-data-attributes.qmd)
- Review [two-table verbs](https://dplyr.tidyverse.org/articles/two-table.html) with dplyr

### Questions

How do you decide the “right” structure for a dataset when it can be organized in more than one tidy way? - Vrinda

Typically, whatever format allows you to efficiently complete the necessary analysis and produce the expected outputs is the "right" structure.

------------------------------------------------------------------------

How much does it matter when making the decision on what function to use when smoothing? - Dillon

Check out the [smoothr](https://strimas.com/smoothr/) documentation for more details.

You can also use `sf::st_simplify()` or `rmapshaper::ms_simplify()` for make less "smooth" lines and polygons.

------------------------------------------------------------------------

Which carrier has the worst average delays? Check [R for Data Science (2e) - Solutions to Exercises](https://mine-cetinkaya-rundel.github.io/r4ds-solutions/) for tips.

```{r}
nycflights13::flights |>
  group_by(carrier) |>
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) |>
  slice_max(order_by = avg_delay, n = 1)
```

::: footer
There are also solutions available for [Geocomputation with R](https://r.geocompx.org/solutions/).
:::

------------------------------------------------------------------------

Can you disentangle the effects of bad airports vs. bad carriers? (via [Solutions Manual: R for Data Science (2e)](https://aditya-dahiya.github.io/RfDS2solutions/))

```{r}
nycflights13::flights |>
  group_by(dest, carrier) |>
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) |>
  # taking the highest average delay flight at each airport
  slice_max(order_by = avg_delay, n = 1) |>
  ungroup() |>
  # for each airline, summarize the number of airports where it is
  # the most delayed airline
  summarise(n = n(), .by = carrier) |>
  slice_head(n = 5) |>
  arrange(desc(n)) |>
  rename(Carrier = carrier, `Number of Airports` = n) |>
  gt::gt()
```

------------------------------------------------------------------------

How do you choose when to use `st_intersection()` vs. `st_join()` when looking at relationships between layers? - Liam

```{r}
nc <- st_read(system.file("shape/nc.shp", package = "sf"))

nc_points <- st_as_sf(st_sample(nc, 250))

st_join(nc_points, nc[1, ])

st_intersection(nc_points, nc[1, ])
```

## Week 5

- Reminder
  - Exercises 2 due Friday, 2/27
  - Exercise 3 due Wednesday, 3/4
- Interesting and Difficult Things!
- Week 5 Questions
- Finish review of [Transforming data with `{dplyr}`](/slides/data-transformation.qmd)
- [Creating and manipulating attributes for spatial data](/slides/spatial-data-attributes.qmd)

### Interesting Things

- "honestly was just surprised how easy it is to perform geometry operations"
- "we can use distance based joins on two datasets that are meaningfully related even when they don’t intersect"

### Difficult Things

- "learning how to do the same thing in multiple different ways"
- "understanding the different binary operators... st_intersects and st_disjoint were intuitive but others like st_covered_by were less clear"
- "finding the time to do the readings"

### Questions

"Have you ever had to use DE-9IM strings, will I ever have to use them, can they be practically effectively used by people who are not deep into the lore??" - Lauren

------------------------------------------------------------------------

"Do unary and binary geometry operations change the input file or allow you to create a unique output file" - Connor

## Week 4

- Updates and reminders
  - Common issues with syntax
  - Exercise how-to with Quarto
  - Exercises 2 and 3 due Friday, 2/27 and Wednesday 3/4 (links coming soon)
  - Office Hours on Friday, 2/20 at 12:00 pm
- Week 4 Questions
- Finish review of data visualization example with ggplot2
- [Transforming data with `{dplyr}`](/slides/data-transformation.qmd)

### Questions

When subseting with the `[` operator, why do you need a comma and space at the end? e.g. `world[world$area_km2 < 10000, ]` —Chase

Use `?[` to take a look at the documentation. When using the `[` operator to subset a data frame (or an `sf` object), the first value is the row index and the second is the column index.

------------------------------------------------------------------------

```{r}
# Select second row
us_states[2, ]

# Select second column
us_states[, 2]
```

------------------------------------------------------------------------

Do you always use `sf` objects when working with spatial data? Or do you switch between spatial and non-spatial formats? —Liam

Yes. Always drop the geometry using `sf::st_drop_geometry()` if you don't need it in your output!

```{r}
#| code-fold: true
#| code-summary: "Summarising a data frame vs. sf object with `bench::mark`"

storms_sf <- sf::st_as_sf(storms, coords = c("long", "lat"), crs = 4326)

comparison <- bench::mark(
  df = storms |>
    dplyr::summarise(
      mean_wind = mean(wind),
      .by = year
    ),
  sf = storms_sf |>
    dplyr::group_by(year) |>
    dplyr::summarise(
      mean_wind = mean(wind)
    ),
  check = FALSE
)
```

------------------------------------------------------------------------

```{r}
#| echo: false
plot(comparison)
```

------------------------------------------------------------------------

Does it matter the order that you specify parameters for a `ggplot`? —Lauren

Consistent [code style](https://r4ds.hadley.nz/workflow-style.html) improves the readability of your code and reduces risk of errors *but* `{ggplot2}` supports a flexible approach.

------------------------------------------------------------------------

This works...

```{r}
ggplot(data = us_states) +
  geom_sf(aes(fill = total_pop_15))
```

------------------------------------------------------------------------

...and this works...

```{r}
ggplot(data = us_states) +
  aes(fill = total_pop_15) +
  geom_sf()
```

------------------------------------------------------------------------

...and this works!

```{r}
ggplot() +
  geom_sf(data = us_states, aes(fill = total_pop_15))
```

------------------------------------------------------------------------

But... this does not work! Do you know why?

```{r}
ggplot() +
  geom_sf(us_states, aes(fill = total_pop_15))
```

### Finish review

Download the example script to review:

```{r}
#| eval: false
usethis::use_github_file(
  repo = "bldgspatialdata/website",
  path = "examples/data-visualization.R"
)
```

## Week 3

- Week 3 Questions
- [Session 2 Quiz](https://docs.google.com/forms/d/e/1FAIpQLScFfOG6aFaMv-pogR2wZHoSYnfhmaemRwHyytWCZPpSJV6dog/viewform?usp=header)
- [Visualizing spatial data with {ggplot2}](/slides/data-visualization.qmd)

### Questions

What are some effective ways to familiarize yourself with the language of different packages without rote memorization of their functions? —Brian

- [Cheatsheets](https://posit.co/resources/cheatsheets/), e.g. [RStudio](https://rstudio.github.io/cheatsheets/html/rstudio-ide.html), [ggplot2](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
- Community resources, e.g. [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
- Naming conventions, e.g. `geom_<type of geometry to show on plot>`

------------------------------------------------------------------------

Is there an easy way to plot summary statistics (e.g. mean, min, max)? —Lauren

```{r}
summary(spData::us_states_df)
```

------------------------------------------------------------------------

Is there an easy way to plot summary statistics (e.g. mean, min, max)? —Lauren

```{r}
skimr::skim(spData::us_states_df)
```

------------------------------------------------------------------------

Why are some CRS values `NA` while some are specific datum (WGS84, etc)? —Nhi

```{r}
x <- st_point(c(1, 0))

st_crs(x)

st_crs(spData::us_states)
```

------------------------------------------------------------------------

Why do some of the exercises start with the library function and not have you initially install a library? —Kyle

```{r}
# install.packages("tidyverse")
```

------------------------------------------------------------------------

What actually is an observation and how is it different than a variable? —Dillon

A variable is something you can *measure*. An observation is a set of *measurements*.

------------------------------------------------------------------------
