---
code-link: true
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
```

## Week 8 Agenda

-   Check-in on difficult + interesting things
-   Feedback on course experience (things to change, add, or keep)
-   Errors of the week
-   **Lecture + practice:** Exploratory data analysis with `{sf}` and the tidyverse
-   Weekly questions

### Check-in

This week's difficult things:

-   Quarto can be confusing
-   Writing functions is hard
-   Naming functions is even harder

This week's interesting things:

-   Functions can be a powerful tool
-   Quarto has an exciting level of flexibility

### Course feedback from you (and me)

Things to change:

-   Start with slides + code demonstration (saving the check-ins + questions for the end)
-   Try a remote session where students can screen share, see and understand each other's work, and receive help on our code
-   Share more feedback on completed exercises\*

### Course feedback from you (and me)

Things to add:

-   Distributing slides in advance of class and spend class time on coding exercises and trouble-shooting
-   Spend more time on how to make maps with `{sf}`, `{ggplot2}`, and R
-   Share exercise solutions for review\*

### Course feedback from you (and me)

Things to keep the same:

-   Exercises that include debugging a code snippet
-   Sharing good questions for weekly check-ins*

### Errors for the week

`IllegalArgumentException: Points of LinearRing do not form a closed linestring`

Take a look at `sf::st_is_valid()` and `sf::st_make_valid()`.

### Errors for the week

`Error in UseMethod("mutate") : no applicable method for 'mutate' applied to an object of class "function"`

Always double-check your inputs and outputs! When an error tells you it has `no applicable method` for a specific input—you may be passing a different object than you think you are. 

### Lecture

🖥️ [Exploratory data analysis](/slides/exploratory-analysis.html)

### Questions on tidy data

**After you tidy data, where or how do you start looking in order to make sense of it?**

### Questions on Quarto

**Why did Quarto render a bunch of figures from the quarto md files of exercises due in previous weeks?**

**How do you duplicate of a Quarto document?**

### Questions on functions

**Is there a way to embed a function within one another?**

Yes! You can define a function within a function like this:

```{r}
example_fn <- function(...) {
  
  embedded_fn <- function(..., collapse = ",") {
    paste0(..., collapse = collapse)
  }
  
  embedded_fn(...)
}
```

------------------------------------------------------------------------

You can also use [anonymous functions and shortcuts](https://adv-r.hadley.nz/functionals.html#purrr-shortcuts) (especially inside of tidyverse functions that translate the "tilde" formula style notation into a function):

```{r}
map_dbl(mtcars, \(x){length(unique(x))})

map_dbl(mtcars, ~ length(unique(.x)))
```

------------------------------------------------------------------------

**Is there any set of naming convention for functions that most people use?**

Use what the function *does* as a guide.

Try an `object_verb()` naming scheme for functions take a common data type or interact with a common API where `object` refers to the data/API and `verb` the primary action. (see the [Function and argument](https://devguide.ropensci.org/building.html#function-and-argument-naming) recommendations in the[ Packaging Guide](https://devguide.ropensci.org/building.html) section of the [rOpenSci Packages](https://devguide.ropensci.org/index.html) book)

For example:

- **stringi**, functions starting with `stri_` manipulate strings (`stri_join()`, `stri_sort()`
- **googlesheets** functions starting with `gs_` are calls to the Google Sheets API `gs_auth()`, `gs_user()`, `gs_download()`

----

**When creating a function is there a way to save it to your R project so if you were to start a new script document you can still use the function?**

------------------------------------------------------------------------

**How do we create a data frame and store the output from a function in a data frame?**

------------------------------------------------------------------------

**What are the advantages of using `return()` over using nested logic (e.g. `if (...) { } else { }`) when writing functions?**

------------------------------------------------------------------------

**How do we determine these thresholds for other variables in different datasets?**

------------------------------------------------------------------------

**If the list of our arguments/inputs are column values from a data frame, how do we call the entire list of our arguments/inputs from that data frame after the function executes?**
