## Week 6: October 4, 2023

## Weekly questions

### Geometric and spatial operations

While the concept of a buffer is rather easy to understand for me in other spatial software such as ArcGIS Pro, it was difficult for me to manipulate in R.

----

```{r}
storms_sf <- storms |> 
  st_as_sf(
    coords = c("long", "lat"),
    crs = 4326
  )

storms_tracks <- storms_sf |> 
  arrange(year, month, day, hour) |> 
  group_by(year, name) |> 
  summarise(
    avg_wind = mean(wind),
    geometry = st_cast(st_union(geometry), to = "LINESTRING")
    )

storms_tracks
```


----

What do I mean by "states that intersect Maryland"?

----

How do you create the map (straight line between MD and CA) using ggplot instead of plot?

```{r}
#| eval: false
plot(
  us_states$geometry,
  main = "States crossed by a straight line from central Maryland to central California"
  )

plot(
  states_crossed$geometry,
  col = "grey",
  add = TRUE
  )

plot(
  md_to_cali,
  color = "black",
  add = TRUE
  )
```

### Assorted questions

How can I streamline my workflow?

### Tidy data questions

----

Are most built in datasets in R already tidy?

----

Can functions from the `{units}` work with any type of variable?

----

When might you use both `tidyr::pivot_longer()` and `tidyr::pivot_wider()`? 

----

What is the purpose of the `.value` parameter used in `names_to` attribute of `pivot_longer()` function?

```{r}
anscombe %>%
  pivot_longer(
    everything(),
    cols_vary = "slowest",
    names_to = c(".value", "set"),
    names_pattern = "(.)(.)"
  )
```


::: notes
".value" indicates that the corresponding component of the column name defines the name of the output column containing the cell values, overriding values_to entirely.
:::

----

How does `tidyr::pivot_wider()` handle cases with conflicting or duplicate values when pivoting?

