## Week 7: October 11, 2023

These past updates are in reverse chronological order.

### Due this week

* Exercise 6 is due this Friday!
* Readings on scripts, functions, and Quarto. 

### Due next week

* Make a mapping function and use it in a Quarto document!

### Your interesting things!

----

The reasoning behind why sometimes you can use pipes and sometimes not finally clicked with me! I had been trying to use pipes for everything and didn't understand why it wasn't working -- getting that down was really helpful.

----

I have found that it super interesting to determine how to tidy data, like taking the time to make logical sense of the data is great but this also leads to what I find most difficult.

----

I really enjoyed the independence of going out and located are own data and being able to manipulate it in R. Feels like the first step into the real world.

----

Joining is just *so much easier* in R as opposed to ArcGIS or QGIS.

----

### Your difficult things

One of the most challenging concepts or topics for me from all the syllabus covered till now is the `st_()` functions.

There are a lot of geometry operations and each has a different set of parameters and features, which are mostly confusing.

### Your questions

#### Trouble-shooting code questions

Why does `plot()` make 6 maps instead of 1?

```{r}
maryland <- us_states |> 
  select(NAME) |> 
  filter(NAME == "Maryland")

states_intersecting_md <- st_filter(
  us_states,
  maryland,
  .predicate = st_intersects
  ) 

plot(states_intersecting_md)
```

----

Why does this not work?

```{r}
ggplot(
  data = states_intersecting_md,
  mapping = aes(x = long, y = lat)
  ) +
geom_sf()
```

----

What is `stringr::str_flatten()` doing?

`str_flatten()` reduces a character vector to a single string.

This is a **summary function** because regardless of the length of the input `x`, it always returns a single string.

```{r}
commas <- function(x) {
  str_flatten(x, collapse = ", ", last = " and ")
}

commas(c("cat", "dog", "pigeon"))
```

----

#### Questions about finding and citing data

Is there a large list or repository of good places to find data or is it more of a treasure hunt?

Here is [a big list of possible data sources](https://bldgspatialdata.github.io/website/resources/data-sources.html) including local and national data sources and a set of R packages focused on data access. Many libraries create similar resources for students and researchers to use and some libraries have geospatial data specialists who can assist you in a search.

There are also a wide variety of scientific data repositories that are typically specific to a domain of research, community of practice, or institution. For example, take a look at this [NIH directory of 130+ data respositories](https://sharing.nih.gov/data-management-and-sharing-policy/sharing-scientific-data/repositories-for-sharing-scientific-data) or the [Environmental Data Initiative (EDI) repository](https://edirepository.org/) used by the long-running Baltimore Ecosystem Study.

---

Are there more sources for *untidy* data? Are there any purposely untidy to be used for tidying practice?

It turns out there is a [package dedicated to sharing a selection of untidy datasets](https://github.com/kwstat/untidydata2)!

`{tidyr}` also comes with a set of tiny practice datasets (`table1`, `table2`, `table3`, `table4a`, `table4b`, `table5`) for practicing these skills. These [slides from Data Science in a Box](https://datasciencebox.org/course-materials/_slides/u2-d09-tidying/u2-d09-tidying#1) or the [accompanying video](https://www.youtube.com/watch?v=x3KM5uxaFdI) are good resources to review the basics of using the `pivot_` and `separate_` function families to change the strcture of your data.

----

What is the best approach for proper citation and credit for data if there is no metadata and it is not clear who created it?

Create a citation with as much information as you have available using a citation manager such as [Zotero](https://www.zotero.org/). But, also reflect: if you don't know who created the data or how they created the data, should you be *using* the data?

#### Questions about data formats

What are the advantages or disadvantages of working with CSV files vs. GeoJSON files vs. other file types?

----

How do I access data using a Feature Service URL?

<https://services1.arcgis.com/UWYHeuuJISiGmgXx/arcgis/rest/services/CustomerServiceRequest_Prod/FeatureServer/0>

#### Questions about working with Quarto


Can use the RStudio debugging tools for code blocks in a `.qmd` file (Quarto document)?

Not really. You could keep your code in one or more R scripts that you can then source into your Quarto document. Or you could extract your code chunks into a R script using the `knitr::purl()` function (for de-bugging purposes only or to use as a source for the document).

----

What is the keyboard shortcut to delete empty code chunks?

It doesn't exist (although it was proposed [via GitHub issue back in November 2021](https://github.com/rstudio/rstudio/issues/10147)). You could probably create one using the [`{shrtcts}` package](https://pkg.garrickadenbuie.com/shrtcts/) or by creating a custom [RStudio Addin](https://rstudio.github.io/rstudioaddins/).
