---
title: "Exercise 02 (Solutions - Incomplete)"
order: 2
status: Available
date-due: 2023-09-11
date-modified: last-modified
---

Incomplete solutions.

## Setup

This exercises uses the `{ggplot2}` and `{dplyr}` packages (both from the [tidyverse family](https://www.tidyverse.org/) of packages) and the `{sf}` package:

```{r}
library(ggplot2)
library(dplyr)
library(sf)
```

For this week's exercise, we are also going to use data from the `{rnaturalearth}` package. Make sure to install those packages and re-start your session if these packages are not installed already:

```{r}
# pak::pkg_install(c("rnaturalearth", "rnaturalearthdata"))
library(rnaturalearth)
```

We are going to use `ne_download()` to download the `countries` dataset and then use `st_centroid()` to make a version of this dataset where the features show the center of each country instead of the boundaries:

```{r}
countries <- ne_download(scale = "medium", type = "countries", returnclass = "sf")

countries <- filter(countries, REGION_UN != "Antarctica")
countries <- st_transform(countries, crs = 3857)

countries_center <- st_centroid(countries)

glimpse(countries)
```

Some of the following exercises don't require a sf object. You can also use the `mpg` or `storms` dataset we looked at during this week's lecture: 

```{r}
glimpse(mpg)

glimpse(storms)
```

One advantage of using `{ggplot2}` over a map-making focused package like `{tmap}` is the wide variety of extension packages created by the large community of users and developers making data visualizations (including maps) with `{ggplot2}`.

We aren't going to use some of these extension packages (and others) but we won't load them yet. Run the following code to install (and don't forget to restart your session afterwards):

```{r}
#| eval: false
pak::pkg_install(c("patchwork", "plotly", "smoothr"))
```

If you finish this exercise but still want more practice, you can download this [RMarkdown document with exercises](https://github.com/thomasp85/ggplot2_workshop/blob/master/exercises.Rmd) shared by Thomas Lin Pedersen for his two-part 2020 online workshop on `{ggplot2}` ([part 1](https://www.youtube.com/watch?v=h29g21z0a68) and [part 2](https://www.youtube.com/watch?v=0m4yywqNPVY) are both available on YouTube).

## Exercises

### Plotting a single variable

Find a discrete variable in countries and then create a plot with `geom_bar()`:

```{r}
ggplot(data = countries) +
  geom_bar(mapping = aes(y = SUBREGION))
```

Next, find a continuous variable and make a plot with `geom_histogram()`:

```{r}
ggplot(data = countries) +
  geom_histogram(mapping = aes(x = POP_EST))
```

Now, let's make a map! Use `countries_center` and `geom_sf()` to make a map with a continuous variable mapped to size:

```{r}
ggplot(data = countries_center) +
  geom_sf(aes(size = POP_EST))
```

----

Next, make a map with `geom_sf()` with one discrete variable mapped to color:

```{r}
ggplot(data = countries) +
  geom_sf(aes(color = REGION_UN))
```

Is that the map you expected? Try it again with the discrete variable mapped to fill:

```{r}
ggplot(data = countries) +
  geom_sf(aes(fill = REGION_UN))
```

----

Now, make a plot using any `geom` function of your choice:

```{r}
ggplot(data = countries) +
  geom_jitter(aes(REGION_UN, POP_EST, color = REGION_UN))
```

Explain in plain language. What does your plot show? ____

### Plotting two variables

For this next section, you can continue to use `countries` as your dataset or load a different dataset using `{rnaturaleath}`. You can see what vector data is available using `rnaturalearth::ne_find_vector_data()` then find the function you need to load the data into a new object:

```{r}
rnaturalearth::ne_find_vector_data()

populated_places <- rnaturalearth::ne_download(type = "populated_places", returnclass = "sf")
```

First, find two continuous variables and create a scatter plot with `geom_point()`:

```{r}
ggplot(data = populated_places) +
  geom_point(aes(MAX_AREAKM, POP_MAX))
```

Next, look in your data for one discrete and one continuous variable then use `aes()` to set those variables for `geom_col()`. The `geom_col()` function is similar to `geom_bar()` but you *must* provide *both* an x *and* a y variable:

```{r}
north_central_america <- filter(countries, SUBREGION %in% c("Northern America", "Central America"))

ggplot(data = north_central_america) +
  geom_col(aes(POP_EST, NAME))
```

Now, use `geom_sf()` mapping your continuous variable to fill or color and your discrete variable to `facet_wrap()`:

```{r}
ggplot(data = countries) +
  geom_sf(aes(fill = POP_EST), color = NA) +
  facet_wrap(~ TYPE)
```

Finally, create a map using a different aesthetic that we haven't tried yet. Options could include `linewidth`, `size`, `alpha`, or `linetype`:

```{r}
ggplot(data = populated_places) +
  geom_sf(aes(size = POP1950))
```

Is this aesthetic mapping an effective way of visualizing the variable? ____

If so, why do you think it works well? If not, why does it *not* work well? ____

### Using scales and colors

 `{ggplot2}` uses naming conventions to organize the scale functions. This isn't the same for every function but they look something like: "scale_<aesthetic>_<color scheme>_<type of data>". So, `scale_fill_viridis_d()` applies the Viridis color scale to a discrete variable mapped to the fill aesthetic.
 
Use the data to create a plot and take a look at the colors set when you use `scale_color_viridis_c()`:

```{r}
ggplot(data = populated_places) +
  geom_point(aes(LATITUDE, LONGITUDE, color = POP2015)) +
  scale_color_viridis_c()
```

The [ColorBrewer](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) scales are designed for use with [thematic maps](https://en.wikipedia.org/wiki/Thematic_map). Use `?scale_color_brewer()` to pull up the documentation for this function and review the information on the type and palette parameters.

Now, map a variable to the `color` aesthetic for `geom_sf()` and assign an appropriate type and palette value:

```{r}
ggplot(data = countries_center) +
  geom_sf(aes(color = REGION_UN)) +
  scale_color_brewer(type = "qual", palette = "Accent")
```

Switching from `color` to `fill`, try it again with a different type and palette value:

```{r}
ggplot(data = countries) +
  geom_sf(aes(fill = INCOME_GRP)) +
  scale_fill_brewer(type = "seq", palette = "GnBu")
```

One last time, but we're using `scale_fill_distiller()`:

```{r}
ggplot(data = countries) +
  geom_sf(aes(fill = GDP_MD)) +
  scale_fill_distiller(type = "seq", palette = "GnBu")
```

Note that this `scale_fill_distiller()` scale only works with continuous values. If you get an error, you may need to map a different variable to `fill`.

### Adding labels, legends, and themes

Set the data for `ggplot()` and then use the `labs()` function to apply a title and caption that make sense:

```{r}
ggplot(data = north_central_america) +
  geom_sf(color = "black", fill = NA) +
  labs(
    title = "North and Central American Countries",
    caption = "Data from the rnaturalearth package"
  )
```

Now, map `fill` to a variable in your data using `aes()` and then use `labs()` to assign a label for fill:

```{r}
ggplot(data = north_central_america) +
  geom_sf(aes(fill = GDP_MD)) +
  labs(
    fill = "GDP"
  )
```

Finally, put all of these elements together with a theme function. `theme_minimal()` and `theme_void()` are good themes to use for maps but you can explore [all of the options in the ggplot2 documentation](https://ggplot2.tidyverse.org/reference/ggtheme.html):

```{r}
ggplot(data = north_central_america) +
  geom_sf(mapping = aes(fill = GDP_MD)) +
  labs(
    title = "North and Central American countries by GDP",
    caption = "Data from the rnaturalearth package",
    fill = "GDP"
  ) +
  theme_void()
```

### Interactive plots with `{plotly}`

We aren't doing much with interactivity in this class (or exercise) but I did want to give you a chance to try it out using the `ggplotly()` function from the `{plotly}` package:

```{r}
p <- ggplot(data = countries) +
  geom_point(aes(POP_EST, GDP_MD, color = REGION_UN))
  
plotly::ggplotly(
  p = p
)
```

A [directory of `{ggplot2}` extensions](https://exts.ggplot2.tidyverse.org/index.html) is available through the tidyverse website if you want to try more tools for animation or interactivity including [{gganimate}](https://gganimate.com/) or [{ggiraph}](https://davidgohel.github.io/ggiraph/).

### Map making with `{ggplot2}`

By default, any map created with `geom_sf()` will show the graticulates on the map and axis labels with the coordinate values. Add data to this map and then hide these graticules by adding `theme_void()`:

```{r}
ggplot(data = countries) +
  geom_sf(color = "black", fill = NA) +
  theme_void()
```

You can also hide or change [graticules](https://en.wikipedia.org/wiki/Graticule_(cartography)) by using `theme()`. Try setting the `panel.grid` argument to `element_blank()` to hide the grid:

```{r}
ggplot(data = countries) +
  geom_sf(color = "black", fill = NA) +
  theme(
    panel.grid = element_blank()
    )
```

Now, try "zooming" into a selected area of your map using the `xlim` and `ylim` arguments for `coord_sf()`:

```{r}
ggplot(data = countries) +
  geom_sf(color = "black", fill = NA) +
  coord_sf(
    xlim = c(-19836524, 20013016),
    ylim = c(2150688, 18394384)
  )
```

If you have difficulty with this one, look back at [our week 2 slides](https://bldgspatialdata.github.io/website/slides/data-visualization.html#/8/6) for an example showing how to use `sf::st_bbox()` to get xmin, xmax, ymin, and ymax values for the xlim and ylim parameter.

----

Using an inset map or "locator map" with a larger area and a zoomed in map showing a featured area is a common cartographic approach. You can use `patchwork::inset_element()` from the `{patchwork}` package to set this up:

```{r}
area_map <- ggplot(data = north_central_america) +
  geom_sf(color = "black", fill = NA) +
  coord_sf(
    xlim = c(-19836524, 20013016),
    ylim = c(2150688, 18394384)
  ) +
  theme_void()

inset_map <- ggplot(data = countries) +
  geom_sf(color = "black", fill = NA) +
  theme_void()

area_map +
  patchwork::inset_element(
    p = inset_map,
    left = 0.1,
    bottom = 0.1,
    top = 0.3,
    right = 0.3
  )
```

Remember, this format of calling functions (`<package name>::<function name>`) is just a shortcut for using functions from packages that *are* installed but not loaded into your environment. If you have any difficulty with this part of the exercise, make sure you have `{patchwork}` installed.

---

There are some cases when you need to modify the geometry of your data as part of the process of making a map. The `st_simplify()` function is one way to do that. Try setting dTolerance to a low value, e.g. `dTolerance = 10`, and run the code block. Then try to run it again with `dTolerance = 100000`.

```{r}
usa <- filter(countries, NAME == "United States of America")

simple_usa <- st_simplify(x = usa, dTolerance = 100000)

ggplot() +
  geom_sf(
    data = usa,
    color = "orange"
    ) +
  geom_sf(
    data = simple_usa,
    color = "purple"
  ) +
  theme_void()
```

What happens when you increase the value of dTolerance? ____

Now, let's try to same thing but smoothing features with `smoothr::smooth()` instead of simplifying with `sf::st_simplify()`. Start by setting smoothness to a small number, `smoothness = 0.5`, and then run again with higher and higher numbers:

```{r}
smooth_usa <- smoothr::smooth(x = usa, method = "ksmooth", smoothness = 100)

ggplot() +
  geom_sf(
    data = usa,
    color = "orange"
    ) +
  geom_sf(
    data = smooth_usa,
    color = "purple",
    fill = NA
  ) +
  theme_void()
```

What happens when you increase the value of smoothness? ____

Check the documentation for `st_simplify()` or `smoothr::smooth()` for more information on how these functions work to modify the geometry.

## Bonus exercise

### Creating maps with `{tmap}`

Pick one of the maps you created in the prior questions of this exercise and create a similar version using the `{tmap}` package.

You can install `{tmap}` the same as any other package:

```{r}
# pak::pkg_install("tmap")
```

Then load the library:

```{r}
library(tmap)
```

And make a map using data from `{rnaturalearth}` or another source of your choice:

```{r}
____
```

What is the same about making a map with `{tmap}` compared to `{ggplot2}`? ____

What is different about making a map with `{tmap}` compared to `{ggplot2}`? ____

Do you have any preference between the two? ____
