---
title: "Exercise 04"
order: 4
date-due: 2023-09-25
date-modified: last-modified
---

::: {.callout-important appearance="minimal" icon="true"}
Exercise due on {{< meta date-due >}}
:::

ℹ️ See [week
{{< meta order >}}](https://bldgspatialdata.github.io/website/weeks/week_0%7B%7B%3C%20meta%20order%20%3E%7D%7D.html)
for related slides and readings

## Overview

This week's exercises are excerpted from [Ch. 3 Attribute data operations](https://r.geocompx.org/attr) in Geocomputation with R. These exercises build on our last exercise using `{dplyr}` and include some of the same skills including:

- Filtering or subsetting rows using attributes

New skills you will practice with this exercise include:

- Working with `unit` class attributes (columns)
- Joining datasets

## Setup

```{r}
library(tidyverse)
library(sf)
```

We are also going to use the `us_states` and `us_states_df` data from the `{spData}` package:

```{r}
library(spData)
```

## Exercises

### Searching for `us_states`

Find all states with the following characteristics (bonus find and plot
them):

Belong to the Midwest region.

```{r}
us_states |> 
  ____
```

Belong to the West region, have an area below 250,000 km2and in 2015 a
population greater than 5,000,000 residents (hint: you may need to use
the function `units::set_units()` or `as.numeric()`).

```{r}
us_states |> 
  ____
```

Belong to the South region, had an area larger than 150,000 km2 or a
total population in 2015 larger than 7,000,000 residents.

```{r}
us_states |> 
  ____
```

What was the total population in 2015 in the `us_states` dataset? What
was the minimum and maximum total population in 2015?

```{r}
us_states |> 
  ____
```

### Comparing and summarizing data

Add variables from `us_states_df` to `us_states`, and create a new
object called `us_states_stats`.

-   What function did you use and why?
-   Which variable is the key in both datasets?
-   What is the class of the new object?

```{r}
us_states_stats <- us_states |> 
  ____
```

`us_states_df` has two more rows than `us_states`. **How can you find
them?**

Hint: try to use the `dplyr::anti_join()` function.

```{r}
____(us_states, us_states_df)
```

**How much has population density changed between 2010 and 2015 in each
state?**

Calculate the change in percentages and map them.

```{r}

```

Calculate the change in the number of residents living below the poverty level between 2010 and 2015 for each state. (Hint: See `?us_states_df` for documentation on the poverty level columns.)

```{r}

```

Bonus: Calculate the change in the percentage of residents living below the poverty level in each state.

```{r}

```


What was the minimum, average and maximum state’s number of people living below the poverty line in 2015 for each region?

```{r}

```

Bonus: What is the region with the largest increase in people living below the poverty line?

```{r}

```


