---
title: "Exercise 05"
order: 5
date-modified: last-modified
---

ℹ️ See [week {{< meta order >}}](/weeks/week_0{{< meta order >}}.qmd) for related slides and readings

## Overview

This week's exercises are excerpted from [Ch. 25 Functions](https://r4ds.hadley.nz/functions.html) in R for Data Science. This exercise is practice for writing three types of functions:

-   Mutate (or vector) functions
-   Data frame functions
-   Plotting functions

## Mutate (or vector) functions

Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need?

```{r}
mean(is.na(x))
mean(is.na(y))
mean(is.na(z))

x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)

round(x / sum(x, na.rm = TRUE) * 100, 1)
round(y / sum(y, na.rm = TRUE) * 100, 1)
round(z / sum(z, na.rm = TRUE) * 100, 1)
```

Given a vector of birthdates, write a function to compute the age in years.

```{r}

```

::: callout-tip
The `{lubridate}` package makes working with dates in R much easier. Take a look at the package [Getting started guide](https://lubridate.tidyverse.org/articles/lubridate.html) or [Ch. 17 Dates and Times](https://r4ds.hadley.nz/datetimes) from R4DS for more information.
:::

## Data frame functions

Using the datasets from `{nycflights13}`, write a function that:

1.  Finds all flights that were cancelled (i.e. `is.na(arr_time)`) or delayed by more than an hour.

```{r}
#| eval: false
flights |> filter_severe()
```

2.  Counts the number of cancelled flights and the number of flights delayed by more than an hour.

```{r}
#| eval: false
flights |> group_by(dest) |> summarize_severe()
```

3.  Finds all flights that were cancelled or delayed by more than a user supplied number of hours:

```{r}
#| eval: false
flights |> filter_severe(hours = 2)
```

4.  Summarizes the weather to compute the minimum, mean, and maximum, of a user supplied variable:

```{r}
#| eval: false
weather |> summarize_weather(temp)
```

5.  Converts the user supplied variable that uses clock time (e.g., `dep_time`, `arr_time`, etc.) into a decimal time (i.e. hours + (minutes / 60)).

```{r}
#| eval: false
flights |> standardize_time(sched_dep_time)
```

::: callout-tip
Question 4 and 5 may require an understanding of [tidy evaluation](https://r4ds.hadley.nz/functions#indirection-and-tidy-evaluation) using the "embracing" syntax. Go back and review our reading if you are not sure how this works!
:::

## Plotting functions

Build up a rich plotting function by incrementally implementing each of the steps below:

1. Draw a map using an `sf` object.

2. Map an variable from the data to a single aesthetic attribute, such as `fill`, `color`, or `size`.

3. Add annotation to the map using `geom_sf_text()` or `geom_sf_label()`. You could add labels to the highest or lowest values or add labels based on a second variable.

4. **Optional:** Apply a spatial or geometric transformation function, e.g. `sf::st_centroid()`, `sf::st_filter()`, or `sf::st_buffer()`, to part or all of the input data before plotting. Consider how spatial transformation can be part of the process of visualizing spatial data.

5. Add a title and labels. Your function should allow a use to customize the labels as needed.

Please give your function an appropriate name and provide one or more examples showing how your function works with different input data sources.

```{r}
plot_function <- function(
  # Function arguments
  ...
) {
  # Body of the function
}
```
